When does vapor pressure deficit drive or reduce evapotranspiration?

### 2018-04-03 doc/paper/vpd_et_paper.tex
readthrough done, now nee dto markup and send to Pierre:

\begin{abstract}
  Increases to vapor pressure deficit (VPD) and atmospheric demand for
  water are expected with rising atmospheric CO$_2$. While increased
  evapotranspiration (ET) in response to increased atmospheric demand
  seems intuitive, plants are capable of reducing ET in response to
  increased VPD by closing their stomata, in an effort to conserve
  water. Here we examine which effect dominates response to increasing
  VPD: atmospheric demand and increases in ET, or plant physiological
  response (stomata closure) and decreases in ET. We use
  Penman-Monteith, combined with semi-empirical optimal stomatal
  regulation theory and underlying water use efficiency, to develop a
  theoretical framework for understanding how ET responds to increases
  in VPD.

  The theory suggests that for most environmental conditions and plant
  types, plant physiological response dominates and ET decreases with
  increasing VPD. Plants that are evolved or bred to prioritize
  primary production over water conservation (e.g. crops) exhibit a
  higher likelihood of atmospheric demand-driven response (ET
  increasing). However for forest, grass, and shrub plant types, ET
  more frequently decreases than increases with rising VPD. Therefore,
  we may expect reductions in ET due to CO$_2$-induced increases in
  VPD. This work serves as an example of the utility of our simplified
  framework for disentangling land-atmosphere feedbacks at multiple
  scales, including the characterization of ET response in an
  atmospherically drier, enriched CO$_2$ world.

@@

The ET response to changes in VPD alters water partitioning between
the soil and atmosphere. If ecosystem plant response reduces ET with atmospheric
drying then soil moisture will be better conserved. This would seem a
sensible evolutionary strategy to cope with aridity. If stomata were
fully passive \citep [similar to soil pores, e.g. ][]{Or_2013},
increased atmospheric aridity would strongly reduce soil moisture
\citep{Berg_2017}. In turn, this would further increase aridity as low
soil moisture levels increase the Bowen ratio, and cause increased
temperature and atmospheric drying \citep[][]{Bouchet_1963,
  Morton_1965, Brutsaert_1999, Ozdogan_2006, Salvucci_2013,
  Gentine_2016, Berg_2016}. This however would not seem to be a
sensible strategy for plants from an evolutionary standpoint. One may
argue that increases in ET with increasing VPD could increase the
liklihood of precipitation \citep[e.g.,][]{Findell_2011}. However,
increases in ET are not a gaurantee
of increased likelhood of precipitation, which also dependes on the
environemtnal conditions \citep[][]{Gentine_2013}. Futhermore,  any
increases in precipitation are likely to be non-local, such that plants giving up
water to the atmosphere are not gauranteed to reap the benefits of
water returned to the atmosphere. Using this logic, from an ecoysystem
evolutionary perspective, water stored in soil is worth much more than
the \textit{chance} of water returned as precipitation.

@@

The objective of the present manuscript is to use reasonable
approximations as a tool to develop intuition for plant response to
atmospheric drying and evaluate the VPD dependence of ET. This
intuition will aid interpretation of observations, full complexity
models, and facilitate the disentanglement of complex land-atmosphere
feedbacks. In the past, similiar approaches were used to understand
interactions between stomatal conductance, evapotranspiration and the
environment \citep[e.g.,][]{Jarvis_1986, Mcnaughton_1991}. However, at
the time researchers' understandng of the form of VPD's effect on
plant physiology was limited, so they could not explore the
sensitivity of ET to VPD, including VPD's effect on stomatal
conductance and plant function.

Recent results have drastically improved our understanding of VPD's
impact on physiology. \citet{MEDLYN_2011} developed a model for
leaf-scale stomatal conductance ($g_s$), including VPD response, by
combining an optimal photosynthesis theory \citep{Farquhar_1980,
  Katul_2010} with an empirical approach, and extended use of this
model to the ecosystem scale in \citet{Medlyn_2017}. Additionally,
\citet{Zhou_2014} demonstrated that a quantity underlying water use
efficiency $\left(uWUE = \frac{GPP\; \sqrt{VPD}}{ET}\right)$ properly
captures a constant relationship between GPP, ET, and VPD over a
diurnal cycle. uWUE is also remarkably well conserved in the growing
season across space and time, within a PFT \citep{Zhou_2015}. These
results provide novell tools to examine how ET varies in response to
VPD, while explicitly including VPD's effect on stoamtal conductance
for the first time. In this manuscript, we leverage these tools to
derive the theoretical response of ET to VPD using a Penman Monteith
framework. Our theory is validated and tested at multiple
eddy-covariance stations spanning various climates and plant
functional types.

\section{Materials and Methods}
\subsection{Data}
\label{data}
We use both meteorological and eddy-covariance data from the
FLUXNET2015 database (data available at \sloppy
https://fluxnet.fluxdata.org/data/fluxnet2015-dataset/ \sloppy),
including all sites with at least four years of data and observations
of the variables described in the methods section (Section
\ref{methods}). Seventy-three sites met these requirements, and were
grouped into nine plant functional types (PFT) according to the
International Geosphere-Biosphere Programme vegetation classification
scheme \citep{Loveland_1999}: cropland (CRO), grass (GRA), deciduous
broadleaf forest (DBF), evergreen broadlead forest (EBF), evergreen
needleleaf forest (ENF), mixed forest (MF), closed shrub (CSH),
savannah (SAV), and woody savannah (WSA) (site locations in Figure
\ref{map_fig}).

@@

\begin{table}
  \caption{Definition of symbols and variables, with citation for how
    values are calculated, if applicable.}
  \label{definitions}
  \centering \small
  \begin{tabular}{l c c c}
    \hline
    Variable & Description & Units & Citation \\
    \hline
    $e_s$  & saturation vapor pressure & Pa  & - \\ 
    $T$  & temperature  & K & - \\
    $P$  & pressure & Pa  & - \\
    $\Delta$  & $\frac{\partial e_s}{\partial T}$ & Pa K$^{-1}$ & - \\
    $R_{net}$  & net radiation at land surface minus ground heat flux & W m$^{-2}$   & - \\
    $g_a$  & aerodynamic conductance & m s$^{-1}$  & \citet{Shuttleworth_2012} \\
    $\rho_a$  & air density & kg m$^{-3}$  & - \\
    $c_p$  & specific heat capacity of air at constant pressure & J K$^{-1}$ kg$^{-1}$ & - \\
    $VPD$  & vapor pressure deficit & Pa  & - \\
    $\gamma$  & psychometric constant & Pa K$^{-1}$   & - \\
    $g_{s-leaf}$  & leaf-scale stomatal conductance & m s$^{-1}$  & \citet{MEDLYN_2011} \\
    $g_{s}$  &  stomatal conductance & m s$^{-1}$
                                   & \citet{Medlyn_2017} \\
    $g_{1-leaf}$  & leaf-scale slope parameter & Pa$^{0.5}$
                                   & \citet{MEDLYN_2011} \\
    $g_{1}$  & ecoysystem-scale slope parameter & Pa$^{0.5}$ & \citet{MEDLYN_2017} \\
    $R$ & universal gas constant & J mol$^{-1}$ K$^{-1}$ & - \\
    $R_{air}$ & gas constant of air & J  K$^{-1}$ kg$^{-1}$ & - \\
    $\sigma$ & uncertainty parameter & -& - \\
    $c_a$ & CO$_2$ concentration & $\mu$ mol CO$_2$ mol$^{-1}$ air& - \\
    $\lambda$ & marginal water cost of leaf carbon & mol H$_2$O mol$^{-1}$ CO$_2$ & - \\
    $\Gamma$ & CO$_2$ compensation point & - & - \\
    $\Gamma^*$ & CO$_2$ compensation point without dark respiration & - & - \\
    \hline
  \end{tabular}
\end{table}

@@

where $\Gamma^*$ is the CO$_2$ compensation point for photosynthesis
(without dark respiration), and $\lambda$ is the marginal water cost
of leaf carbon
($\frac{\partial \; \text{transpiration}}{\partial \; A}$). So,
$g_{1-leaf}$ is a leaf-scale term reflecting the tradeoff of water for
carbon uptake. The higher $g_{1-leaf}$, the more open the stomata and
the more they release water in exchange for carbon.

@@

where GPP is the ecoysystem scale gross primary production, and $g_1$
is an ecosystem scale analogue to $g_{1-leaf}$. We solve for $g_1$
following \citet{Medlyn_2017} (Equation 5), and take the median $g_1$
value to be representative of each PFT (Table \ref{pft}).

@@

While Equation 3 can be used in PM (Equation 1), it will make
analytical work with the function intractable because $GPP$ is
functionally related to ET itself. Additionally, a perturbation to VPD
should induce a physiological plant response that will alter GPP and
cause an indirect change in stomatal conductance, in addition to the
direct effect of VPD in Equation \ref{medlyn}. Therefore, in order to
derive the response of ET to VPD, we must account for the functional
relationship between GPP, ET, and VPD, and its effect on stomatal
conductance. We can use aforementioned semi-empirical results of
\citet{Zhou_2015} as a tool to approach this
problem. \citet{Zhou_2015}, showed that the underlying Water Use
Efficiency (uWUE):

@@

This derivation relied either implicitly or explicitly on several
assumptions. First, the underlying theory used to derive Equations
\ref{medlyn}, \ref{leaf_medlyn} rely on a implicit assumptions that
$ET=g_s \cdot VPD$; physically this implies that leaves are perfectly
coupled to the atmosphere. In reality, for some conditions and plant
types the leaves can become decoupled from the boundary layer
\citep{De_2017, Medlyn_2017}. Given our focus on the growing season
which is usually characterized by relatively high insolation, we would
expect the surface to be well coupled. Additionally, by following
\citet{Medlyn_2017}'s method for calcualting eco-system scale $g_1$,
we have stronger confidence that surface coupling is represented
within our framework, even if it is not represented in the underlying
theory used to develop $g_s$. Additionally, the formulation of $uWUE$
in \citet{Zhou_2014} and \citet{Zhou_2015} as well as our use of
\citet{Medlyn_2017}'s stomatal conductance model assumes that that
direct soil evaporation contributions to ET remian relatively
small. This should be more true druing the growing season than other
times; however some plant types allow for systematically larger
contributions of ``E'' in ET, particularly those with less dense
canopies and smaller aboslute amounts of transpiration. We therefore
might expect that the theory will be most applicable to forest PFTs,
which will be most strongly coupled to the boundary layer due to
larger surface roughness, and will also generally have the highest
ratios of transpiration to evaporation. Finally, for the goal of
developing PFT-wide intuition, we assume that $g_1$ and $uWUE$ are
constant in space and time. Both quantities have been shown to be well
conserved over space and time within a PFT relative to inter-PFT
variability. However, we would expect them to vary somehwat with
environmental conditions including soil water content temperature, and
with intra-PFT plant-specific characteristics like wood density for
tree PFTs \citep{Lin_2015}. Given these strong assumptions made with
the goal of understanding broad, leading order plant behaviors, we
take the extra care of including and analyzing uncertainty (Section
\ref{testing}) and examing sensitivity to the exponent of the VPD
response derived using assumptions of $ET=g_s \cdot VPD$ (Section
\ref{functional_form}).

We note one final comment on our derivation which is relevant for
drought indices. If we approximate $c_a$ at a global mean CO$_2$
concentration, then the RHS of Equation \ref{et} is fully defined
using commonly available weather station data and the constants
published here. This makes Equation \ref{et} a useful alternative to
PET in drought indices and hydrometeorological analysis for vegetated
surfaces. Equation \ref{et} better reflects the physics of water
exchange at the land surface and would only require fitting of uWUE
and $g_1$, and can also account for changes in CO$_2$ concentration
(assuming some relationship between $\lambda$ and [CO$_2$]), contrary
to typical drought indices such as the Palmer Drought Severity Index
(PDSI) \citep{Swann_2016}.

@@

By construction, the variability in the $\sigma$ term (Equation
\ref{sigma}) contains all model and observational uncertainties. For
an observation that perfectly matches our model and constant uWUE
assumption, $\sigma$ will be one. Therefore, for our assumptions and
framework to be reasonable $\sigma$ should be close to 1. An
additional concern is that $\sigma$ may in fact be correlated with
$VPD$, in which case the dependence would need to be accounted for
when taking the derivative. Fortunately, there is a very weak
dependence of $\sigma$ on VPD in their joint distrubution, and
$\sigma$ is indeed close to unity i.e. $O(1)$ (Figure
\ref{joint_vpd_sigma}). Given this weak dependence and the
distribution of $\sigma$ we have confidence in our model framework and
the data quality.

@@

This effect of uWUE and $g1$ on the sign term is most apparent by
comparing two extreme PFTs: water intensive crops (CRO) and water
conservative closed shrub (CSH). CRO has higher slope parameter and a
slightyly higher uWUE ($g_1 = 140.7$ Pa$^{1/2}$; $2.85$ $\mu$-mol [C]
Pa$^{0.5}$ J$^{-1}$ [ET]) compared to CSH ($g1 = 75.1 \, Pa^{1/2}$,
$uWUE=2.82$ $\mu$-mol [C] Pa$^{0.5}$ J$^{-1}$ [ET]). These differences
in PFT parameters cause opposite ET responses to changes in VPD
between CRO and CSH. ET theoretically always decreases with increasing
VPD for the more water conservative CSH, while ET frequently increases
with increasing VPD for the more water intensive CRO (Figure
\ref{idealized_sign}). CRO evolved or were bred to prioritize GPP and
yield and are thus not water conservative. They are very willing to
trade water for photosynthesis and productivity, despite changes in
VPD, while CSH are very unwilling to trade water for more
photosynthesis.

@@

To summarize our theoretical insights (Figure \ref{idealized_sign} and
Table \ref{vpd_crit}), CROs are the least water conservative and have
the strongest overall tendency to increase ET with increasing VPD,
while CSH are the most water conservative and have the strongest
tendency to decrease ET with increasing VPD, as well as the strongest
VPD dependence of response. Figure \ref{idealized_sign} clearly shows,
according to our theory, that for all PFTs except for crops there is
frequent occurrence of a negative (plant dominating) ET response to
increases in VPD. Therefore, plants are able in most atmospheric
conditions to reduce ET in response to increased VPD and thus to
reduce water loss. To better illustrate this, the ranges of observed
environmental VPDs at the FLUXNET sites are plotted parallel to the
x-axis. For CSH, VPD is always less than VPD$_{crit}$ (off scale) so
that the plant response dominates in typical environmental conditions,
emphasizing the water conservative strategy of those plants. For CRO
on the other hand, VPD is higher than VPD$_{crit}$ for more than 50\%
of observations, emphasizing that those plants operate with an
aggressive water usage strategy, are water intensive and were actually
engineered for photosynthesis rather than water saving. For DBF, EBF,
MF, GRA, and SAV more than half of the observed VPD are less than
VPD$_{crit}$, i.e. in conditions where plant response dominates. SAV
has a more water conservative response than the forest, grass, and
crop plan types, but still respondes by increasing ET with increasing
VPD for about a quarter of observations, due to the high aridity (VPD)
of the SAV ecoclimate. It is also important to note that for all PFTs,
even when atmospheric demand dominates, ET response to VPD is still
far more negative than it would be for potential evaporation
$\partial PET/\partial VPD$, i.e. atmospheric demand only, emphasizing
that there is still a strong regulation of evaporative flux by stomata
and though the plant xylem. The sign term in the PET case would just
be a constant ($\frac{c_p}{R_{air}} \approx 3.5$), which is far larger
than any part of the curves for any PFT. This highlights the
deficiencies of PET's ability to capture land response to changes in
atmospheric dryness (VPD). Plants are always regulating water exchange
from the land surface, even when they reach the limits of they ability
to do so.

@@

The control of the scaling term variability between PFTs by
aerodynamic conductance is confirmed by data (Figure
\ref{scale_vary}). Differences between PFT are almost entirely due to
differences in aerodynamic conductance, rather than differences in
observed temperature ranges. The scaling term for the tree PFTs (DBF,
EBF, ENF, MF) is generally about double the scaling terms for other
PFTs which have lower surface roughness and generally smaller
aerodynamic conductance (GRA, CSH, CRO). The savannah (WSA, SAV) PFT's
scaling is somewhere betwen GRA, CSH, and CRO, and DBF, EBF, ENF, and
MF, due to higher variability and surface roughness.

@@

Regarding the frequency of negative and positive ET response, all PFTs
exhibit a decreasing ET response with increasing VPD (physiologically
controlled, water conservative response) for the majority of
observations. The more water conservative PFTs generally exhibit
higher frequency of negative ET response, espcially when one factors
in the distribution of environmental VPD (e.g. SAV and WSA grow in
more arid climates, Figure \ref{idealized_sign}). In general the bulk
statistics match our theoretical expectations well, with the caveat
that inclusion of uncertainty shifts crops towards a slightly more
negative response, and shifts many of the other PFTs, which still
exhibit a high frequency of negative ET reponse, towards more frequent
occurence of postive ET response than the theory and Figure
\ref{idealized_sign} would suggest. The bulk statistics motivate a
more thorough examination of the structure of uncertainty and more
sophisticated validation of our theory's performance against
observations.

\subsection{Validation of theory at eddy-covariance sites}
\label{testing}
We now compare more sophisticated distributions of the observed
response to our simplified theory (Section \ref{sign_term}). The
observed distribution of the sign term, as compared to what the theory
would predict, is provided in Figure \ref{test_sign}. Our goal was to
capture the leading order behavior of the ET dependence on VPD. Given
the assumptions we have made, and the uncertainties of flux tower
observations themselves, we expect a relatively large amount of noise
when reproducing the derivatives of ET. However, the data largely
reproduces our theoretical analysis.

@@

While CSH has a much different functional form of the sign term than
DBF and MF, CSH observations also match our theory to leading order,
albiet with a bit more variability as a function of VPD. Again, the
VPD dependence mostly determined from the slope parameter ($g_1$)
closely matches the medians in the observation bins. The
VPD-independent, strongly negative response is also captured. For CSH,
there is rare occurrence of observed positive ET response with VPD
($\approx 6$\%), even with uncertainty, so the sign of the
observations almost always matches the sign of the theory, which
states that ET response should always be negative.

>>>>

\begin{abstract}
  Increasing vapor pressure deficit increases atmospheric demand for
  water, and vapor pressure deficit is expected to rise with
  increasing CO$_2$. While increased evapotranspiration (ET) in
  response to increased atmospheric demand seems intuitive, plants are
  capable of reducing ET in response to increased VPD by closing their
  stomata, in an effort to conserve water. Here we examine which
  effect dominates response to increasing VPD: atmospheric demand and
  increases in ET, or plant physiological response (stomata closure)
  and decreases in ET. We use Penman-Monteith, combined with
  semi-empirical optimal stomatal regulation theory and underlying
  water use efficiency, to develop a theoretical framework for
  understanding how ET responds to increases in VPD.

  The theory suggests that for most environmental conditions and plant
  types, plant physiological response dominates and ET decreases with
  increasing VPD. Plants that are evolved or bred to prioritize
  primary production over water conservation (e.g. crops) exhibit a
  higher likelihood of atmospheric demand-driven response (ET
  increasing). However for forest, grass, savannah, and shrub plant
  types, ET more frequently decreases than increases with rising
  VPD. This work serves as an example of the utility of
  our simplified framework for disentangling land-atmosphere feedbacks
  at multiple scales, including the characterization of ET response in
  an atmospherically drier, enriched CO$_2$ world.

@@

The ET response to changes in VPD alters water partitioning between
the soil and atmosphere. If ecosystem plant response reduces ET with
atmospheric drying then soil moisture will be better conserved. This
would seem a sensible evolutionary strategy to cope with aridity. If
stomata were fully passive \citep [similar to soil pores,
e.g. ][]{Or_2013}, increased atmospheric aridity would strongly reduce
soil moisture \citep{Berg_2017}. In turn, this would further increase
aridity as low soil moisture levels increase the Bowen ratio, and
cause increased temperature and atmospheric drying
\citep[][]{Bouchet_1963, Morton_1965, Brutsaert_1999, Ozdogan_2006,
  Salvucci_2013, Gentine_2016, Berg_2016}. This however would not seem
to be a sensible strategy for plants from an evolutionary
standpoint. As a counterpoint, one may argue that increases in ET with
increasing VPD could increase the likelihood of precipitation
\citep[e.g.,][]{Findell_2011}. However, increases in ET do not always
guarantee an increased likelihood of precipitation, which depending on
environmental conditions could cause a decrease in the likelihood
of precipitation \citep[][]{Gentine_2013}. Furthermore, any increases
in precipitation are likely to be non-local, such that plants giving
up water to the atmosphere are not guaranteed to reap the benefits of
water returned from the atmosphere to the soil. Using this logic, from
an ecosystem evolutionary perspective, water stored in soil is worth
much more than the chance of water returned as precipitation.

@@

The objective of the present manuscript is to use reasonable
approximations as a tool to develop intuition for plant response to
atmospheric drying and evaluate the VPD dependence of ET. This
intuition will aid interpretation of observations, full complexity
models, and facilitate the disentanglement of complex land-atmosphere
feedbacks. In the past, similar approaches were used to understand
interactions between stomatal conductance, evapotranspiration and the
environment \citep[e.g.,][]{Jarvis_1986, Mcnaughton_1991}. However, at
the time researchers' understanding of the form of VPD's effect on
plant physiology was limited, so they could not explore the
sensitivity of ET to VPD, including VPD's effect on stomatal
conductance and plant function.

Recent results have drastically improved our understanding of VPD's
impact on physiology. \citet{MEDLYN_2011} developed a model for
leaf-scale stomatal conductance ($g_s$), including VPD response, by
combining an optimal photosynthesis theory \citep{Farquhar_1980,
  Katul_2010} with an empirical approach, and extended use of this
model to the ecosystem scale in \citet{Medlyn_2017}. Additionally,
\citet{Zhou_2014} demonstrated that a quantity underlying water use
efficiency $\left(uWUE = \frac{GPP\; \sqrt{VPD}}{ET}\right)$ properly
captures a constant relationship between GPP, ET, and VPD over a
diurnal cycle. uWUE is also remarkably well conserved in the growing
season across space and time, within a PFT \citep{Zhou_2015}. These
results provide novel tools to examine how ET varies in response to
VPD, while explicitly including VPD's effect on stomatal conductance
for the first time. In this manuscript, we leverage these tools to
derive the theoretical response of ET to VPD using a Penman Monteith
framework. Our theory is validated and tested at multiple
eddy-covariance stations spanning various climates and plant
functional types.

\section{Materials and Methods}
\subsection{Data}
\label{data}
We use both meteorological and eddy-covariance data from the
FLUXNET2015 database (data available at \sloppy
https://fluxnet.fluxdata.org/data/fluxnet2015-dataset/ \sloppy),
including all sites with at least four years of data and observations
of the variables described in the methods section (Section
\ref{methods}). Seventy-three sites met these requirements, and were
grouped into nine plant functional types (PFT) according to the
International Geosphere-Biosphere Programme vegetation classification
scheme \citep{Loveland_1999}: cropland (CRO), grass (GRA), deciduous
broadleaf forest (DBF), evergreen broadleaf forest (EBF), evergreen
needleleaf forest (ENF), mixed forest (MF), closed shrub (CSH),
savannah (SAV), and woody savannah (WSA) (site locations in Figure
\ref{map_fig}).

@@

\begin{table}
  \caption{Definition of symbols and variables, with citation for how
    values are calculated, if applicable.}
  \label{definitions}
  \centering \small
  \begin{tabular}{l c c c}
    \hline
    Variable & Description & Units & Citation \\
    \hline
    $e_s$  & saturation vapor pressure & Pa  & - \\ 
    $T$  & temperature  & K & - \\
    $P$  & pressure & Pa  & - \\
    $\Delta$  & $\frac{\partial e_s}{\partial T}$ & Pa K$^{-1}$ & - \\
    $R_{net}$  & net radiation at land surface minus ground heat flux & W m$^{-2}$   & - \\
    $g_a$  & aerodynamic conductance & m s$^{-1}$  & \citet{Shuttleworth_2012} \\
    $\rho_a$  & air density & kg m$^{-3}$  & - \\
    $c_p$  & specific heat capacity of air at constant pressure & J K$^{-1}$ kg$^{-1}$ & - \\
    $VPD$  & vapor pressure deficit & Pa  & - \\
    $\gamma$  & psychometric constant & Pa K$^{-1}$   & - \\
    $g_{s-leaf}$  & leaf-scale stomatal conductance & m s$^{-1}$  & \citet{MEDLYN_2011} \\
    $g_{s}$  &  stomatal conductance & m s$^{-1}$
                                   & \citet{Medlyn_2017} \\
    $g_{1-leaf}$  & leaf-scale slope parameter & Pa$^{0.5}$
                                   & \citet{MEDLYN_2011} \\
    $g_{1}$  & ecosystem-scale slope parameter & Pa$^{0.5}$ & \citet{MEDLYN_2017} \\
    $R$ & universal gas constant & J mol$^{-1}$ K$^{-1}$ & - \\
    $R_{air}$ & gas constant of air & J  K$^{-1}$ kg$^{-1}$ & - \\
    $\sigma$ & uncertainty parameter & -& - \\
    $c_a$ & CO$_2$ concentration & $\mu$ mol CO$_2$ mol$^{-1}$ air& - \\
    $\lambda$ & marginal water cost of leaf carbon & mol H$_2$O mol$^{-1}$ CO$_2$ & - \\
    $\Gamma$ & CO$_2$ compensation point & - & - \\
    $\Gamma^*$ & CO$_2$ compensation point without dark respiration & - & - \\
    \hline
  \end{tabular}
\end{table}

@@

where $\Gamma^*$ is the CO$_2$ compensation point for photosynthesis
(without dark respiration), and $\lambda$ is the marginal water cost
of leaf carbon
($\frac{\partial \; \text{transpiration}}{\partial \; A}$). So,
$g_{1-leaf}$ is a leaf-scale term reflecting the trade-off of water for
carbon uptake. The higher $g_{1-leaf}$, the more open the stomata and
the more they release water in exchange for carbon.

@@

where GPP is the ecosystem scale gross primary production, and $g_1$
is an ecosystem scale analogue to $g_{1-leaf}$. We solve for $g_1$
following \citet{Medlyn_2017} (Equation 5), and take the median $g_1$
value to be representative of each PFT (Table \ref{pft}).

@@

While Equation 3 can be used in PM (Equation 1), it will make
analytical work with the function intractable because $GPP$ is
functionally related to ET itself. Additionally, a perturbation to VPD
should induce a physiological plant response that will alter GPP and
cause an indirect change in stomatal conductance, in addition to the
direct effect of VPD in Equation \ref{medlyn}. Therefore, in order to
derive the response of ET to VPD, we must account for the functional
relationship between GPP, ET, and VPD, and its effect on stomatal
conductance. We can use aforementioned semi-empirical results of
\citet{Zhou_2015} as a tool to approach this
problem. \citet{Zhou_2015}, showed that  underlying Water Use
Efficiency (uWUE):

@@

This derivation relied either implicitly or explicitly on several
assumptions. First, the underlying theory used to derive Equations
\ref{leaf_medlyn} and \ref{medlyn} rely on an implicit assumption that
$ET=g_s \cdot VPD$; physically this implies that leaves are perfectly
coupled to the atmosphere. In reality, for some conditions and plant
types the leaves can become decoupled from the boundary layer
\citep{De_2017, Medlyn_2017}. Given our focus on the growing season,
which is usually characterized by relatively high insolation inducing
instability and convective boundary layers, we would
expect the surface to be generally well coupled. Additionally, by following
\citet{Medlyn_2017}'s method for calculating ecosystem scale $g_1$
using PM,
we have stronger confidence that surface coupling is represented
within our framework, even if it is not represented in the underlying
theory used to develop stomatal conductance ($g_s$). An additional
assumption in the formulation of $uWUE$
\citep{Zhou_2014, Zhou_2015} and
\citet{Medlyn_2017}'s stomatal conductance model is that
direct soil evaporation (E) contributions to ET remain
small relative to transpiration (T). This should be more true during the growing season. However, some plant types allow for systematically larger
contributions of ``E'' in ET, particularly those with less dense
canopies and smaller absolute amounts of transpiration. We therefore
might expect that the theory will be most applicable to forest PFTs,
which will be most strongly coupled to the boundary layer due to
larger surface roughness, and will also generally have the highest
ratios of transpiration to evaporation. Finally, for the goal of
developing PFT-wide intuition, we assume that $g_1$ and $uWUE$ are
constant in space and time. Both quantities have been shown to be well
conserved over space and time within a PFT relative to inter-PFT
variability. However, we would expect them to vary somewhat with
environmental conditions including soil water content, temperature, and
with intra-PFT plant-specific characteristics like wood density for
tree PFTs \citep{Lin_2015}. Given these strong assumptions made with
the goal of understanding broad, leading order plant behaviors, we
take the extra care of including and analyzing uncertainty (Section
\ref{testing}) and examining sensitivity to the exponent of the VPD
response, which is derived using assumptions that $ET=g_s \cdot VPD$ (Section
\ref{functional_form}).

We note one final comment on our derivation which is relevant for
drought indices. If we approximate $c_a$ at a global mean CO$_2$
concentration, then the RHS of Equation \ref{et} is fully defined
using commonly available weather station data and the constants
published here. This makes Equation \ref{et} a useful alternative to
PET in drought indices and hydrometeorological analysis for vegetated
surfaces. Equation \ref{et} better reflects the physics of water
exchange at the land surface and would only require fitting of uWUE
and $g_1$, and can also account for changes in CO$_2$ concentration
(assuming some relationship between $\lambda$ and [CO$_2$]), contrary
to typical drought indices such as the Palmer Drought Severity Index
(PDSI) \citep{Swann_2016}. Lastly, all code and data used in this
analysis, including those used to generate the figures and tables, are
publicly available at https://github.com/massma/climate\_et.

@@

By construction, the variability in the $\sigma$ term (Equation
\ref{sigma}) contains all model and observational uncertainties. For
an observation that perfectly matches our model and constant uWUE
assumption, $\sigma$ will be one. Therefore, for our assumptions and
framework to be reasonable $\sigma$ should be close to 1. An
additional concern is that $\sigma$ may in fact be correlated with
$VPD$, in which case the dependence would need to be accounted for
when taking the derivative. Fortunately, there is a very weak
dependence of $\sigma$ on VPD in their joint distribution, and
$\sigma$ is indeed close to unity i.e. $O(1)$ (Figure
\ref{joint_vpd_sigma}). Given this weak dependence and the
distribution of $\sigma$ we have confidence in our model framework and
the data quality.

@@

This effect of uWUE and $g1$ on the sign term is most apparent by
comparing two extreme PFTs: water intensive crops (CRO) and water
conservative closed shrub (CSH). CRO has higher slope parameter and a
slightly higher uWUE ($g_1 = 140.7$ Pa$^{1/2}$; $2.85$ $\mu$-mol [C]
Pa$^{0.5}$ J$^{-1}$ [ET]) compared to CSH ($g1 = 75.1 \, Pa^{1/2}$,
$uWUE=2.82$ $\mu$-mol [C] Pa$^{0.5}$ J$^{-1}$ [ET]). These differences
in PFT parameters cause opposite ET responses to changes in VPD
between CRO and CSH. ET theoretically always decreases with increasing
VPD for the more water conservative CSH, while ET frequently increases
with increasing VPD for the more water intensive CRO (Figure
\ref{idealized_sign}). CROs evolved or were bred to prioritize GPP and
yield and are thus not water conservative. They are very willing to
trade water for photosynthesis and productivity, despite changes in
VPD, while CSH are very unwilling to trade water for more
photosynthesis.

@@

To summarize our theoretical insights (Figure \ref{idealized_sign} and
Table \ref{vpd_crit}), CROs are the least water conservative and have
the strongest overall tendency to increase ET with increasing VPD,
while CSH are the most water conservative and have the strongest
tendency to decrease ET with increasing VPD, as well as the strongest
VPD dependence of response. Figure \ref{idealized_sign} clearly shows,
according to our theory, that for all PFTs except for crops there is
frequent occurrence of a negative (plant dominating) ET response to
increases in VPD. Therefore, plants are able in most atmospheric
conditions to reduce ET in response to increased VPD and thus to
reduce water loss. To better illustrate this, the ranges of observed
environmental VPDs at the FLUXNET sites are plotted parallel to the
x-axis. For CSH, VPD is always less than VPD$_{crit}$ (off scale) so
that the plant response dominates in typical environmental conditions,
emphasizing the water conservative strategy of those plants. For CRO
on the other hand, VPD is higher than VPD$_{crit}$ for more than 50\%
of observations, emphasizing that those plants operate with an
aggressive water usage strategy, are water intensive and were actually
engineered for photosynthesis rather than water saving. For DBF, EBF,
MF, GRA, and SAV more than half of the observed VPD are less than
VPD$_{crit}$, i.e. in conditions where plant response dominates. SAV
has a more water conservative response than the forest, grass, and
crop plan types, but still responds by increasing ET with increasing
VPD for about a quarter of observations, due to the high aridity (VPD)
of the SAV ecoclimate. It is also important to note that for all PFTs,
even when atmospheric demand dominates, ET response to VPD is still
far more negative than it would be for potential evaporation
$\partial PET/\partial VPD$, i.e. atmospheric demand only, emphasizing
that there is still a strong regulation of evaporative flux by stomata
and though the plant xylem. The sign term in the PET case would just
be a constant ($\frac{c_p}{R_{air}} \approx 3.5$), which is far larger
than any part of the curves for any PFT. This highlights the
deficiencies of PET's ability to capture land response to changes in
atmospheric dryness (VPD). Plants are always regulating water exchange
from the land surface, even when they reach the limits of they ability
to do so.

@@

The control of the scaling term variability between PFTs by
aerodynamic conductance is confirmed by data (Figure
\ref{scale_vary}). Differences between PFT are almost entirely due to
differences in aerodynamic conductance, rather than differences in
observed temperature ranges. The scaling term for the tree PFTs (DBF,
EBF, ENF, MF) is generally about double the scaling terms for other
PFTs which have lower surface roughness and generally smaller
aerodynamic conductance (GRA, CSH, CRO). The savannah (WSA, SAV) PFT's
scaling is somewhere between GRA, CSH, and CRO, and DBF, EBF, ENF, and
MF, due to higher variability and surface roughness.

@@

Regarding the frequency of negative and positive ET response, all PFTs
exhibit a decreasing ET response with increasing VPD (physiologically
controlled, water conservative response) for the majority of
observations. The more water conservative PFTs generally exhibit
higher frequency of negative ET response, especially when one factors
in the distribution of environmental VPD (e.g. SAV and WSA grow in
more arid climates, Figure \ref{idealized_sign}). In general the bulk
statistics match our theoretical expectations well, with the caveat
that inclusion of uncertainty shifts crops towards a slightly more
negative response, and shifts many of the other PFTs, which still
exhibit a high frequency of negative ET response, towards more frequent
occurrence of positive ET response than the theory and Figure
\ref{idealized_sign} would suggest. The bulk statistics motivate a
more thorough examination of the structure of uncertainty and more
sophisticated validation of our theory's performance against
observations.

\subsection{Validation of theory at eddy-covariance sites}
\label{testing}
We now compare more sophisticated distributions of the observed
response to our simplified theory (Section \ref{sign_term}). The
observed distribution of the sign term, as compared to what the theory
would predict, is provided in Figure \ref{test_sign}. Our goal was to
capture the leading order behavior of the ET dependence on VPD. Given
the assumptions we made, and the uncertainties of flux tower
observations themselves, we expect a relatively large amount of noise
when reproducing the derivatives of ET. However, the data largely
reproduces our theoretical analysis.

@@

While CSH has a much different functional form of the sign term than
DBF and MF, CSH observations also match our theory to leading order,
albeit with a bit more variability as a function of VPD. Again, the
VPD dependence mostly determined from the slope parameter ($g_1$)
closely matches the medians in the observation bins. The
VPD-independent, strongly negative response is also captured. For CSH,
there is rare occurrence of observed positive ET response with VPD
($\approx 6$\%), even with uncertainty, so the sign of the
observations almost always matches the sign of the theory, which
states that ET response should always be negative.
